Hierarchical wrinkling in a confined permeable biogel

### 2015-02-12 fig_plis/article_plis.tex
Modifs de Thomas: - ajoute abstract et conclusion - fait passer le détail de la littérature des plis de l'intro à "Models and scaling laws" - pattern formation -> system design. Explication en profondeur de la non-monotonicité de la solubilité de la caséine et ses conséquences. - dynamics of the wrinkling process: meilleure description de fig:sideview - utilise directement lambda et non lambda* - refait passer tab:data en Supplementary - Remise en ordre des Method summary - Titration - Sample characteristics pour introduire tab:data - Supplementary Movies

\begin{abstract}
\seb{\it Ajouter l'abstract.}
\end{abstract}

@@

There are many ways and reasons for a film to wrinkle. An elastic film buckles due to excess area with respect to its boundaries. Wrinkling is a buckling hindered by a substrate: the mismatch of elastic properties between the film and the substrate selects the preferred wrinkling wavelength~\cite{Gough1940, Bijlaard1946}. On the engineering side the last two decades have seen a bloom of methods to obtain well-controlled patterns via linear~\cite{Bowden1998, Genzer2006, Hu1998, Kim2010, Vandeparre2011, Li2013} or nonlinear wrinkling~\cite{Efimenko2005, Guvendiren2010, Kim2011a, Brau2011}. Such patterns can be triggered by temperature dilation, swelling or the removal of pre-strain. On the biology side wrinkling-controlled morphogenesis is ubiquitous: ageing and the loss of elastic fibres makes our skin wrinkles~\cite{Bissett1987, Genzer2006}, difference in growth rates between the gut tube and its dorsal anchoring is responsible for the vilification of guts~\cite{Savin2011, Ciarletta2014, Shyer2013}, localized cell death in biofilms focuses mechanical forces and initiates 3D labyrinth pattern~\cite{Trejo2013, Asally2012}. On the physics side analogues of wrinkling have been found in situations where an effective substrate comes from the resistance to uniaxial stretching~\cite{Cerda2003}, from gravity~\cite{Kolinski2009, Vella2009, Pineirua2013, Lucantonio2013}, from boundaries~\cite{Vandeparre2011a, Li2013} or from adhesion~\cite{Vella2009a}. Dealing only with (semi)infinite substrates, Biot generalised the elasticity of both the film and the substrate to viscoelasticity~\cite{Biot1957} or poroelasticity~\cite{Biot1964}, leading to time-dependent kinetic wavelength selection. Stability analysis of a film lying on a thin viscous substrate was studied only recently~\cite{Huang2002}. Since in practice gravity sets the wavelength in such situation~\cite{Smoluchowski1910}, the wrinkling on a confined viscous substrate has not been studied experimentally.

Yet, in a biological context, thin tissues, e.g. epithelimum or endothelium, are naturally confined and immersed in a nearly buoyant medium, e.g. lymph, blood or mucus, a practical situation where a thin viscous substrate would set the wrinkling wavelength. Moreover, most biological films are porous, and little is known on the interplay between permeability and wrinkling~\cite{Ma2004,Longley2013}. Therefore benchmark experiments that explore the possibility of wrinkling in confined porous soft materials immersed in a buoyancy-matched viscous medium are in line to get fundamental insights into such a biologically relevant situation.

@@

The main goal of the present study is thus to identify the selection mechanism of the wavelength \seb{and to discuss its implication for practical situations (?)}. In the following we first describe the physical chemistry that allows us to form such patterns. Then we  model the situation and identify two possible selection mechanisms, namely porous Darcy flow or viscous Poiseuille flow. Comparing our experimental data to these two models we finally show that a combination of both Darcy and Poiseuille mechanisms accurately describes our data. This combined model pins down the dynamical origin of the constraints exerted on the gel and nicely predicts the wrinkling wavelength.
%Poiseuille~\cite{Poiseuille1842} flows of the solvent above and below the gel film for long wavelengths and a new mechanism, Darcy~\cite{Darcy1856} flows of the solvent through the gel film for short wavelengths.

\section*{Pattern formation}

@@

The slow acidification of a protein (sodium caseinate) solution down to the isoelectric pH induces gelation~\cite{Bremer1989} (Fig.~\ref{fig:acidification}a). Since caseins strongly adsorb to interfaces, the resulting gel sticks to the container walls, allowing rheological measurements (Fig.~\ref{fig:acidification}b). Further decrease of the pH beyond the isoelectric point makes the casein aggregates swell (Supplementary Fig.~\ref{fig:nonmonotonic}), thus weakening the gel~\cite{Braga2006}. In a microscopy cell of slit geometry coated so that the casein cannot adsorb on the top wall, this microscopic nonmonotonic behaviour translates into little variation of the pore size $\xi$ but to solvent expulsion out of the gel phase (syn\ae{}resis) followed by global swelling (Fig.~\ref{fig:acidification}c and Supplementary Fig.~\ref{fig:nonmonotonic}).

Figure~\ref{fig:sideview}a shows confocal measurements in the case where both top and bottom walls have been coated. Starting from a homogeneous solution (first panel Fig.~\ref{fig:sideview}a where black margins correspond to the glass walls), the gel forms and simultaneously expels solvent on both sides. Since two interfaces clearly separate a protein-rich phase and a protein-poor phase, we model the resulting situation as a porous elastic film of thickness $h$ sandwiched between two viscous layers of respective thicknesses $H_1$ and $H_2$ in a cell of thickness $e=H_1+h+H_2$ (Fig.~\ref{fig:sideview}b). Wavelength being much larger than thickness, we can assume small strain in the film and lubrication in the layers. Relative increase in protein concentration during syn\ae{}resis can be deduced from the brighter fluorescence in the second panel of Fig.~\ref{fig:sideview}a. Extracting the top and bottom surface of the gel phase allows for quantitative measurements of the relative volume (Fig.~\ref{fig:sideview}c) and excess surface area of the gel film (Fig.~\ref{fig:sideview}d) (see Supplementary Methods). This confirms that the instability is due to swelling.

Note that adhesion to side walls is necessary to prevent total syn\ae{}resis and thus allow subsequent planar stresses to develop while swelling: the only situation were wrinkling was observed is when both top and bottom walls are coated. We also confirmed that at any given point in time and space the thickness $h$ of the film of gel is homogeneous. The contrast in transmission microscopy (Fig.~\ref{fig:dynamics}a) is therefore not due to thickness inhomogeneities but to altitude gradients. Indeed the length of the optical path along the vertical direction $z$ through the gel phase is larger where the gel makes a slope compared to a flat situation, as indicated by the yellow vertical lines in Fig.~\ref{fig:sideview}b.

@@

Longitudinally constrained swelling of the gel film is the origin and driving force of the buckling instability. The gel area becomes in excess compared to the available area, thus the gel must fold. However, this does not explains how a wavelength is selected.

The characteristic length $\lambda^*$ of a wrinkling pattern is set by the competition between the flexural modulus of the film $B=E/(1-\nu^2)$ ($E$ Young modulus, $\nu$ Poisson ratio) and a transverse load $\sigma_\perp$~\cite{Vella2009,Kolinski2009}: 
\begin{equation}
\lambda^* \sim h \left(\frac{B}{\sigma_\perp}\right)^{1/3}.
\label{eq:lstar}
\end{equation}
For small deflections the actual wavelength $\lambda$ is related to $\lambda^*$ and to the wrinkle amplitude $A$ as $\lambda \sim A^{1/4} \lambda^{*3/4}$.

% Pour Mathieu : ajouter quelque part des applications numériques pour les trois longueurs d'ondes gravité, Darcy, Poiseuille
We identify three possible candidates for the transverse load $\sigma_\perp$: the weight or buoyancy of the gel, the porous flow through the gel or the viscous flows in the surrounding water layers. We confirmed that the former plays no significant role by successfully repeating our experiment in a cell held vertically, i.e. with the weight acting longitudinally rather than transversely.
%Since the gel is almost buoyant with the surrounding medium, we expect that the weight plays no significant role, with an elasto-gravitational length at least centimetric. We confirmed this insight by successfully repeating our experiment in a cell held vertically, i.e. with the weight acting longitudinally rather than transversely.

@@

In a first limit where the film of gel would be sitting without sticking to the bottom wall of the cell ($H_1\rightarrow0$), creating a 2D blister requires flowing solvent of viscosity $\eta$ through the gel of permeability $\alpha$. Darcy law~\cite{Darcy1856} relates the pressure gradient, estimated as $\sigma_\perp/h$, to the volume flux per unit area $v = \frac{\alpha}{\eta}\frac{\sigma_\perp}{h} = \frac{dA}{dt}$. Injecting this expression in Eq.~(\ref{eq:lstar}) yields the wavelength of the Darcy mode as
\begin{equation}
\lambda_D \sim h^{1/2} \alpha^{1/4} \Upsilon^{1/4}
\end{equation}
where $\Upsilon = B\tau/\eta$ is a dimensionless viscoelastic factor quantifying the relative stiffness of the film and the solvent at a characteristic time $\tau$. This expression emphasizes the kinetic nature of the wavelength selection through Darcy flow. However we note that the wavelength of a generation of the pattern does not evolve with time (see Fig.~\ref{fig:dynamics}a and Sup. Fig.~\ref{fig:growth}) once the gel is touching both walls. This arrest may be due to bending rigidity~\cite{LeGoff2014}. The relevant $\tau$ corresponds to the times it takes for the amplitude to saturate (see Supplementary Method for the precise definition).

The opposite limiting case is when the gel film lies in the middle of the cell ($H_1=H_2$) and impermeable ($\alpha\rightarrow 0$). There destabilisation over a wavelength $\lambda$ creates a lubrication ($H\ll\lambda$) flow in the viscous layers. By symmetry the transverse load across the gel $\sigma_\perp$ is also the pressure difference over the wavelength. Using a Poiseuille profile for the flow, $\frac{dA}{dt} \sim \sigma_\perp \frac{H^3}{\eta\lambda^2}$~\cite{Poiseuille1842}. Injecting this expression in Eq.~(\ref{eq:lstar}) yields the wavelength of the Poiseuille mode as
\begin{equation}
\lambda_P \sim (hH)^{1/2} \Upsilon^{1/6}
\end{equation}
where we recognize the scaling for the wrinkles of an elastic film on a thin ($H\ll\lambda$) elastic substrate~\cite{Cerda2003} where $\Upsilon$ would be the ratio between the flexural modulus of the film and that of the substrate. Indeed within the lubrication approximation or at low Reynolds numbers one can consider a viscous film as elastic with effective Young modulus $3\eta/\tau$~\cite{Biot1957,Boudaoud2001}.

@@

To sum up, we expect two regimes of viscosity-dominated wrinkling: for $H\ll H^*$ the main dissipation mechanism would be the flow through the porous film, whereas for $H\gg H^*$ the flow in the viscous layers would dominate at set the wavelength.

@@

For a typical gel sheet (Table~\ref{tab:data}b) the buoyancy after syn\ae{}resis is estimated as $\Delta\rho=\SI{130}{\kilo\gram\per\cubic\metre}$ leading to a gravity stress $\sigma_\perp^G \approx \SI{0.04}{\pascal}$. This is much lower than typical Darcy or Poiseuille stresses $\sigma_\perp^D \approx \sigma_\perp^P \approx \SI{0.15}{\pascal}$.

Darcy and Poiseuille models yield wavelengths that are of the same order in most of the cases under study. Thus only systematic measurements of the experimental $\lambda$ together with estimates for all the parameters involved in either model shall allow one to discriminate between them (see Supplementary Method for details on such measurements). As summarized in Table~\ref{tab:data} we systematically varied (i) the thickness of the cell, impacting $h$ and the maximal amplitude, (ii) the gel composition and thus both its stiffness and its permeability, (iii) the viscosity of the solvent by adding glycerol, which also impacted the gel stiffness, permeability and $\tau$. Finally, the only parameter that we had no control upon was the initial altitude $H$ of the gel before wrinkling that varied randomly. However we were able to obtain more data at small $H/H^*$ ratios by considering not only the primary patterns but also the secondary ones. Since the saturation of the primary pattern flattens the gel on the top and bottom walls, the secondary blisters always appear in a Darcy situation ($H\ll H^*$). Here the wavelength associated to the secondary blister is taken as half of that of the primary pattern as shown in Supplementary Fig.~\ref{fig:generations}.

@@

\seb{\it Ajouter une conclusion.}
%optical elements ?
%wallpaper ?
%role of gravity

\section*{Methods summary}
To prevent adsorption of caseins to surfaces, microscope slides and cover slips were cleaned and subsequently coated with a silane agent (3-(trimethoxysilyl) propyl methacrylate, Sigma Aldrich) to which a polyacrylamide brush was polymerized from methacrylate groups.
Sodium caseinate (Firmenich) was labelled with Dylight~550 NHS ESTER (Thermo Scientific). Excess dye was removed by centrifugation. Unlabelled sodium caseinate powder was dissolved in deionised water (or water--glycerol mixture). To induce gelation glucono-$\delta$-lactone (GDL) in powder (Firmenich) was added to the solution. \SI{45}{\micro\litre} of this solution was immediately mixed with \SI{5}{\micro\litre} of labelled caseinate solution, stirred and injected in the microscopy cell, immediately sealed using ultraviolet-cured glue (Norland Optical) and placed under the microscope before gelation could take place. 
3D data were collected on a Zeiss LSM510 confocal microscope, using \SI{532}{\nano\meter} laser excitation. We used either a 10x (air) lens to measure the geometry of wrinkling (see Supplementary Material) or a 63x (oil) lens to observe the microstructure of the gel. Larger scope pictures, obtained either by stitching fluorescent microscopy images (Nikon Eclipse Ti) or by transmitted/reflected light macroscope (Nikon SMZ745T/Leica DMS1000) were used to measure the wavelength of the primary pattern (Supplementary Figures~\ref{fig:patterns} and \ref{fig:generations}). Secondary pattern wavelength is set to half of the primary wavelength, according to Supplementary Fig.~\ref{fig:generations}.
Data shown in Fig.~\ref{fig:acidification} a-b were obtained with a SevenCompact pH-meter (Mettler Toledo) and an MCR 301 rheometer (Anton Paar).
Permeability is measured using a protocol found in \cite{VanDijk1986} (see Supplementary Method and Supplementary Fig.~\ref{fig:permeability}).

>>>>

\begin{abstract}
The slow acidification of milk proteins such as casein is well known to result in the formation of gels – this is indeed one of the key steps to make cheese or yogurt. Here, using the casein proteins as  building blocks and the acidifier  Glucono delta-lactone, we design model thin bio-gel layers that wrinkles in cascade. Using a combination of titration, rheology, light microscopy and confocal microscopy, we demonstrate that, during the acidification, the gels shrink and swell creating constrains that induce Darcy and Poiseuille flows and wrinkles the gel. We have develop a hydrodynamic model that predict correctly the wrinkling wavelength. Our results should be universal for proteins as it is based on pH induced charge stabilasion/destablisation and control of boundary conditons. As such it could set a benchmark to gain fundamental insights into wrinkled biological tissues, to texture food or to design surfaces for optical puposes.
\end{abstract}

@@

%There are many ways and reasons for a film to wrinkle. An elastic film buckles due to excess area with respect to its boundaries. Wrinkling is a buckling hindered by a substrate: the mismatch of elastic properties between the film and the substrate selects the preferred wrinkling wavelength~\cite{Gough1940, Bijlaard1946}. On the engineering side the last two decades have seen a bloom of methods to obtain well-controlled patterns via linear~\cite{Bowden1998, Genzer2006, Hu1998, Kim2010, Vandeparre2011, Li2013} or nonlinear wrinkling~\cite{Efimenko2005, Guvendiren2010, Kim2011a, Brau2011}. Such patterns can be triggered by temperature dilation, swelling or the removal of pre-strain. On the biology side wrinkling-controlled morphogenesis is ubiquitous: ageing and the loss of elastic fibres makes our skin wrinkles~\cite{Bissett1987, Genzer2006}, difference in growth rates between the gut tube and its dorsal anchoring is responsible for the vilification of guts~\cite{Savin2011, Ciarletta2014, Shyer2013}, localized cell death in biofilms focuses mechanical forces and initiates 3D labyrinth pattern~\cite{Trejo2013, Asally2012}. On the physics side analogues of wrinkling have been found in situations where an effective substrate comes from the resistance to uniaxial stretching~\cite{Cerda2003}, from gravity~\cite{Kolinski2009, Vella2009, Pineirua2013, Lucantonio2013}, from boundaries~\cite{Vandeparre2011a, Li2013} or from adhesion~\cite{Vella2009a}. Dealing only with (semi)infinite substrates, Biot generalised the elasticity of both the film and the substrate to viscoelasticity~\cite{Biot1957} or poroelasticity~\cite{Biot1964}, leading to time-dependent kinetic wavelength selection. Stability analysis of a film lying on a thin viscous substrate was studied only recently~\cite{Huang2002}. Since in practice gravity sets the wavelength in such situation~\cite{Smoluchowski1910}, the wrinkling on a confined viscous substrate has not been studied experimentally.

There are many ways and reasons for a film to wrinkle. An elastic film buckles due to excess area with respect to its boundaries and wrinkles when buckling is hindered by a substrate. The selection of the preferred wrinkling wavelength is peculiar to the situation and is related to the mismatch of elastic properties between the surface and the substrate~\cite{Gough1940, Bijlaard1946} XXX.  On the biology side wrinkling-controlled morphogenesis is ubiquitous. Ageing and the loss of elastic fibres makes our skin wrinkles~\cite{Bissett1987, Genzer2006}. Difference in growth rates between the gut tube and its dorsal anchoring is responsible for the vilification of guts~\cite{Savin2011, Ciarletta2014, Shyer2013}. Localized cell death in biofilms focuses mechanical forces and initiates 3D labyrinth pattern~\cite{Trejo2013, Asally2012}. On the physics side the last two decades have seen a bloom of methods to obtain well-controlled patterns via linear~\cite{Bowden1998, Genzer2006, Hu1998, Kim2010, Vandeparre2011, Li2013} or nonlinear behaviour~\cite{Efimenko2005, Guvendiren2010, Kim2011a, Brau2011}. Such patterns can be triggered by temperature dilation refXXX, swelling refXXX or the removal of pre-strain refXXX.

%ph swelling: hu_1998_Polymer gels with engineered environmentally and kim_2010_Dynamic display of biomolecular patterns
%compressive stress: kim_2011_Hierarchical folding of elastic membranes under

Yet,  benchmark experiments that explore the possibility of wrinkling in confined porous soft materials immersed in a buoyancy-matched viscous medium are in line, not only from a physics point of view, but also to get fundamental insights into biologically relevant situation. Indeed, stability analysis of a film lying on a thin viscous substrate was studied only recently~\cite{Huang2002} and remains theoretical because, on a free interface, gravity dominates over viscous substrate. In a biological context, thin tissues, e.g. epithelimum or endothelium, are naturally confined and immersed in a nearly buoyant medium, e.g. lymph, blood or mucus, a practical situation where a thin viscous substrate would set the wrinkling wavelength. Moreover, most biological films are porous, and little is known on the interplay between permeability and wrinkling~\cite{Ma2004,Longley2013}.

@@

The main goal of the present study is thus to identify the selection mechanism of the wavelength. In the following we first describe the physical chemistry that allows us to form such patterns. Then we  model the situation and identify two possible selection mechanisms, namely porous Darcy flow or viscous Poiseuille flow. Comparing our experimental data to these two models we finally show that a combination of both Darcy and Poiseuille mechanisms accurately describes our data. This combined model pins down the dynamical origin of the constraints exerted on the gel and nicely predicts the wrinkling wavelength.
%Poiseuille~\cite{Poiseuille1842} flows of the solvent above and below the gel film for long wavelengths and a new mechanism, Darcy~\cite{Darcy1856} flows of the solvent through the gel film for short wavelengths.

\section*{System design}

@@

To form wrinkled patterns such as the one oberved in figure 1XXX, ones needs to take advantage of the pH sensitivity of casein to control the route to gelation, to fine tune the gel properties and the boundary conditions.

the stability of casein dispersion strongly depend on the pH of the solution. Supplementary movie 3XXX shows the titration of a casein dispertion at 1\%w by 1 molar HCl. At pH=7, the casein is well dispersed then it floculate between pH= XXX to XXX and finaly at pH=XXX the solution become clear again. However as one increases the casein concentration up to 4\%w, at low pH the casein redispersion is not total, Supplementary movie 4XXX. The state diagram of casein as a function of the pH  is related to charge stabilization. At neutral pH casein are charged negatively and repel each other. As the pH decreases to the isoelectric point of the caseine (pH=4.5), caseine molecules loose their negative net charge and tend to stick to each other [ref XXX casein stability]. Beyond the isoelectric point casein acquire a positive net charge and therefore partialy re-disolve. In figure 2cXXX, we quantify this effect and measure with UV absorption spectroscopie the amount of free casein within the 4\%w dispersion as a function of pH (see Supplementary Methods). Starting at pH=6.8 all the casein molecules are well dispersed and the free casein fraction is 1. Upon decreasing the pH of the solution to the isolectric point the free casein fraction drop to almost 0, almost all the casein molecule are aggregated. Further decrease of the pH of the solution lead to a partial redisolution of the casein agregate. At pH=2 the fraction of free casein is XXX.

%On the Stability of Casein Micelles, Pieter Walstra J. of dairy science 1990

To form a space spaning gel, floculation need to be prevented. We use of Glucono delta-lactone (GDL) a molecule  that slowly hydrlolyses into an acid and thus lowers slowly, continuously and homogeneously the pH of the medium, fig.2aXXX. This GDL controled acidification permits the formation of a bulk gel phase. We monitor the GDL induced gelation with a rheometer in cone plate geometry. As casein strongly adsorbs to surfaces, the strain and stress are well transmitted to the casein dispersion and we can measure the elatic and viscous modulus of the casein dispersion as a function of time. the mechanical properties of the gel are closely related to the fraction of free casein. First, the dispersion solidifies then reaches its maximum elasticity (500 Pa) at the isoelectric point and finaly the elastic modulus decreases down to 100 Pa.

To gain further insight into the mechanical properties of the gel we turn to micro structural observations under confocal microscopy. As shown in supplementary fig. 1XXX the gel consist in a space spaning fractal network with a curtoff length scale xi. xi corresponds to the maximum pore size of the network. This model fits the structure at all times but its parameters vary non monotoneously during GDL acidification. As shown in Fig. 2cXXX, up to the isolectric point xi increases up to a maximum length of 7 $\mu$m due to the solvent expulsion out of the casein network [refXXX]. This stiffen the arm of the network and is responsible for the increase of elasticity. Beyond the isoelectric point xi decreases down to 4$\mu$m. Casein regain solubility [refXXX], the strand network swell, the pore shrinks and the gel weakens.

As seen on the confocal measurements, the gel is porous and is permeable to the solvant.  This protertie is determinant as it allow the gel to absord and expel the solvant and therefore change its volume when sticky boundary are turned off. To quantify this effect, we have embeded the casein gel inside a capillary and deeped verticaly the capilary in an aqueous buffer [refXXX]. In the supplementary figure XXX, we observe that the aqueous buffer goes through the gel and rise in the tube due to capillary forces. Exploitation of this experiment give the permeability of the gel: 67000 nm$^2$.

Boundary conditions is the final parameter that is essential to form wrinkled patterns. The evolution of xi when sticky boundaries are on reflects the inner constrains of the gel:  xi increases, the stand of the nework become thinner and the gel wants to retract; then xi decreases, the stands of network swell and the gel wants to expend. When the casein adhesion to the upper boundary is turned off, the gel still not wrinkles because it remains attached to the bottom wall, but it detach from the top wall. In this situation we can monitor xi, fig. XXX. xi remains mostly constant and the major effect is repported on the volume occupied by the gel as will be repported below.

%We coat the glass slides with acrylamide brushes to prevent the adhesion of the casein. In fig.2c, when the top slide of the optical cell is coated with acrylamide brushes, we then observe that the casein network detaches from the top slide. xi does not peak at the isoelectric point but remains constant

%Boundary conditions is the final parameter that is essential to form wrinkled patterns. Once the space spanning network is formed and sticks to the container, protein attraction can only proceed by local shrinking of the protein agregates, leading to larger pores.

%The difference in value of xi between the fully un-coated  and the half-coated optical cell demonstrate that the gel is subject to constrains as the pH decreases. The relaxation of thoses constrain is key to the wrinkling formation when both the top and bottom slide are coated with acrylamide brushes.

%on a change les condition daccroche ce aui permet au contrainte de se relacher

%The slow acidification of a protein (sodium caseinate) solution down to the isoelectric pH induces gelation~\cite{Bremer1989} (Fig.~\ref{fig:acidification}a). Since caseins strongly adsorb to interfaces, the resulting gel sticks to the container walls, allowing rheological measurements (Fig.~\ref{fig:acidification}b). Further decrease of the pH beyond the isoelectric point makes the casein aggregates swell (Supplementary Fig.~\ref{fig:nonmonotonic}), thus weakening the gel~\cite{Braga2006}. In a microscopy cell of slit geometry coated so that the casein cannot adsorb on the top wall, this microscopic nonmonotonic behaviour translates into little variation of the pore size $\xi$ but to solvent expulsion out of the gel phase (syn\ae{}resis) followed by global swelling (Fig.~\ref{fig:acidification}c and Supplementary Fig.~\ref{fig:nonmonotonic}).

\section*{Dynamics of the wrinkling process}

Figure~\ref{fig:sideview}a shows confocal measurements in the case where both top and bottom walls have been coated, a configuration favorable to wrinkling. Up to 10 min the dispersion is homogeneous and fills the entire slit. The black margins correspond to the glass walls. Around 23 min, the gel forms and simultaneously becomes thiner and more concentrated - the fluorescence signal which is proportional to the casein concentration is brighter compared to the situation at 10min. This effect is the fingerprint of syn\ae{}resis: as gelation proceeds, the solvent is expeled on both the upper and lower sides. We describe sample as a porous elastic film of thickness $h$ sandwiched between two viscous layers of respective thicknesses $H_1$ and $H_2$ in a cell of thickness $e=H_1+h+H_2$ (Fig.~\ref{fig:sideview}b). Extracting the top and bottom surface of the gel phase allows for quantitative measurements of the relative volume (Fig.~\ref{fig:sideview}c) and excess surface area of the gel film (Fig.~\ref{fig:sideview}d) (see Supplementary Methods). As expected from the system design, the evolution of the volume is non-monotoneous. After reaching a minimum around 35\% of its initial value at 23 min, the volume increases. Shortly after,  at 36 min, the planar film destabilizes and starts to wrinkle in the $z$-direction creating excess area. This confirms that the instability is due to swelling. Around 38 min, the amplitude of the wrinkles reach the thickness of the slit. The size of the primary pattern is now settled Supplementary Fig. \ref{fig:patterns, fig:growth}. However swelling continues, the gel flattens on both top and bottom walls (44 min) and finally buckles back (53 min) to form the secondary pattern (1h21). When the excess volume saturates, the dynamics stops. The amount of GDL controls the ability of the gel to swell: at low GDL concentrations we only observe the primary pattern, at high GDL concentrations we have observe the formation of higher order patterns.

Note that adhesion to side walls is necessary to prevent total syn\ae{}resis and thus allow subsequent planar stresses to develop while swelling: the only situation were wrinkling was observed is when both top and bottom walls are coated with acrylamide brushes. We also confirmed that at any given point in time and space the thickness $h$ of the film of gel is homogeneous. The contrast in transmission microscopy (Fig.~\ref{fig:dynamics}a) is therefore not due to thickness inhomogeneities but to altitude gradients. Indeed the length of the optical path along the vertical direction $z$ through the gel phase is larger where the gel makes a slope compared to a flat situation, as indicated by the yellow vertical lines in Fig.~\ref{fig:sideview}b.

@@

%Longitudinally constrained swelling of the gel film is the origin and driving force of the buckling instability. The gel area becomes in excess compared to the available area, thus the gel must fold. However, this does not explains how a wavelength is selected.

The determination of the characteristic length $\lambda$ of a wrinkling pattern is an old problem. Large deformations of thin sheets is governed by F\"{o}ppl-von Karman equations  refLaudauXXX,  essentially impossible to solve in analytical form. However using scaling and asymptotic arguments one can obtain the dominant wavelength of the wrinking patern and isolate the physically meaningful ingredients. An elastic film buckles due to excess area with respect to its boundaries. Wrinkling is a buckling hindered by a substrate: the mismatch of elastic properties between the film and the substrate selects the preferred wrinkling wavelength~\cite{Gough1940, Bijlaard1946}. $\lambda$ is then set by the competition between the flexural modulus of the film $B=E/(1-\nu^2)$ ($E$ Young modulus, $\nu$ Poisson ratio) and a transverse load $\sigma_\perp$. For small deflections of  amplitude $A$, $\lambda$ scales as~\cite{Vella2009,Kolinski2009}: 
%
\begin{equation}
\lambda \sim \left( A h^3 \frac{B}{\sigma_{\perp}} \right)^{1/4}.
\label{eq:lstar}
\end{equation}

%L. Landau and E.M. Lifshitz, Theory of Elasticity (Pergamon, New York, 1986), 3rd ed.

This framework has been successfully applied to situations where $\sigma_{\perp}$ comes from the resistance of an elastic substrate ~\cite{Gough1940, Bijlaard1946}. Dealing only with (semi)infinite substrates, Biot generalised the elasticity of both the film and the substrate to viscoelasticity~\cite{Biot1957} or poroelasticity~\cite{Biot1964}, leading to time-dependent kinetic wavelength selection. Futhermore, \ref{eq:lstar} remains valids even when $\sigma_{\perp}$ comes from the resistance to uniaxial stretching~\cite{Cerda2003}, from gravity~\cite{Smoluchowski1910, Kolinski2009, Vella2009, Pineirua2013, Lucantonio2013}, from boundaries~\cite{Vandeparre2011a, Li2013} or from adhesion~\cite{Vella2009a}.

In our situation we now have to determine $\sigma_\perp$. We identify three possible candidates for the transverse load $\sigma_\perp$: the weight or buoyancy of the gel, the porous flow through the gel or the viscous flows in the surrounding water layers. We confirmed that gravity plays no significant role by successfully repeating the wrinkling patern in a cell held vertically, i.e. with the weight acting longitudinally rather than transversely.

@@

We consider two limit scenario that leads to wrinkling parterns. In a first limit, the film of gel sits without sticking to the bottom wall of the cell ($H_1\rightarrow0$). Creating a 2D blister requires flowing solvent of viscosity $\eta$ through the gel of permeability $\alpha$. Darcy law~\cite{Darcy1856} relates the pressure gradient, estimated as $\sigma_{\perp}/h$, to the volume flux per unit area $v = \frac{\alpha}{\eta}\frac{\sigma_{\perp}}{h} = \frac{dA}{dt}$. Injecting this expression in Eq.~(\ref{eq:lstar}) yields the wavelength of the Darcy mode as
\begin{equation}
\lambda_D \sim h^{1/2} \alpha^{1/4} \Upsilon^{1/4}
\end{equation}
where $\Upsilon = B\tau/\eta$ is a dimensionless viscoelastic factor quantifying the relative stiffness of the film and the solvent at a characteristic time $\tau$. This expression emphasizes the kinetic nature of the wavelength selection through Darcy flow. However we note that the wavelength of a generation of the pattern does not evolve with time (see Fig.~\ref{fig:dynamics}a and Sup. Fig.~\ref{fig:growth}) once the gel is touching both walls. This arrest may be due to bending rigidity~\cite{LeGoff2014}. Here the relevant time scale $\tau$ corresponds to the times it takes for the wrinkle amplitude to saturate (see Supplementary Method for the precise definition).

The opposite limiting case is when the gel film lies in the middle of the cell ($H_1=H_2$) and impermeable ($\alpha\rightarrow 0$). There destabilisation over a wavelength $\lambda$ creates a lubrication ($H\ll\lambda$) flow in the viscous layers. By symmetry the transverse load across the gel $\sigma_{\perp}$ is also the pressure difference over the wavelength. Using a Poiseuille profile for the flow, $\frac{dA}{dt} \sim \sigma_{\perp} \frac{H^3}{\eta\lambda^2}$~\cite{Poiseuille1842}. Injecting this expression in Eq.~(\ref{eq:lstar}) yields the wavelength of the Poiseuille mode as
\begin{equation}
\lambda_P \sim (hH)^{1/2} \Upsilon^{1/6}
\end{equation}
where we recognize the scaling for the wrinkles of an elastic film on a thin ($H\ll\lambda$) elastic substrate~\cite{Cerda2003} where $\Upsilon$ would be the ratio between the flexural modulus of the film and that of the substrate. Indeed within the lubrication approximation or at low Reynolds numbers one can consider a viscous film as elastic with effective Young modulus $3\eta/\tau$~\cite{Biot1957,Boudaoud2001}.

@@

To sum up, we expect two regimes of viscosity-dominated wrinkling: for $H\ll H^*$ the main dissipation mechanism would be the flow through the porous film, whereas for $H\gg H^*$ the flow in the viscous layers would dominate and set the wavelength.

@@

For a typical gel sheet (Supplementary Table~\ref{tab:data}b) the buoyancy after syn\ae{}resis is estimated as $\Delta\rho=\SI{130}{\kilo\gram\per\cubic\metre}$ leading to a gravity stress $\sigma_\perp^G \approx \SI{0.04}{\pascal}$. This is much lower than typical Darcy or Poiseuille stresses $\sigma_\perp^D \approx \sigma_\perp^P \approx \SI{0.15}{\pascal}$ and therefore gravity is dismissed.

Darcy and Poiseuille models yield wavelengths that are of the same order in most of the cases under study. Thus only systematic measurements of the experimental $\lambda$ together with estimates for all the parameters involved in either model shall allow one to discriminate between them (see Supplementary Method for details on such measurements). As summarized in Supplementary Table~\ref{tab:data} we systematically varied (i) the thickness of the cell, impacting $h$ and the maximal amplitude, (ii) the gel composition and thus both its stiffness and its permeability, (iii) the viscosity of the solvent by adding glycerol, which also impacted the gel stiffness, permeability and $\tau$. Finally, the only parameter that we had no control upon was the initial altitude $H$ of the gel before wrinkling that varied randomly. However we were able to obtain more data at small $H/H^*$ ratios by considering not only the primary patterns but also the secondary ones. Since the saturation of the primary pattern flattens the gel on the top and bottom walls, the secondary blisters always appear in a Darcy situation ($H\ll H^*$). Here the wavelength associated to the secondary blister is taken as half of that of the primary pattern as shown in Supplementary Fig.~\ref{fig:generations}.

@@

\section*{conclusion}

In conclusion we have taken advantage of the properties of the protein casein and carefully ingineered the route to form permeable bio-gels that, in slit geometry,  produce original cascade of wrinkling patterns  when casein adhesion to the slit walls is turned off.  We developed a model that captures the dynamical origin of the constrains exerted on the gel and predict the wrinkling wavelength: Poiseuille flows of the solvent above and below the gel layer for long wavelengths and a new mechanism, Darcy flows of the solvent through the gel layer for short wavelengths. Such a model experiment, because it relies on pH induced charge stabilisation/destablisation, should be generalisable to different sort of proteins. As such it illustrate daily life problems such as wrinkles encountered while putting up wall paper and could set a benchmark to explore potential applications in food sience refXXX or in optical devices such as diffraction gratins or fresnel lenses refXXX, and the possibility of wrinkling in confined porous soft materials immersed in a buoyancy-matched viscous medium such as biological tissues refXXX.

%role of gravity

\section*{Methods summary}

Sodium caseinate (Firmenich)  is disolved in deionised water by stiring for 20 min  at room temperature.  The mixing of the caseine dispersion with glucono-$\delta$-lactone (GDL) in powder (Firmenich) marks the begining of the experiments. We tipically have about 10 min to pipette the dispersion in the slit before gelation starts. Sodium caseinate (Firmenich) is labelled with Dylight~550 NHS ESTER (Thermo Scientific). Excess dye is removed by centrifugation. For confocal microscopy measurements, we add 10\% of fluorescently labelled caseines to the caseine/GDL dispersion while it is fluid.
